---
author: "Eden fagerstrom"
date: '2022-05-10'
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### Abstract 
Shooting is the one of the most important skills in basketball as it is essential 
in order to score points for your team which will ultimately determine the outcome
of the game. This report examines the effects of distance and ball type on making 
a successful shot into a basketball hoop. The 3 ball types tested in this experiment
are Basketball, Soccerball and Netball. This experiment helped produce a model to 
predict making a successful shot based on distance. The Statistical analysis showed 
Ball type did not effect the success of making a shot although basketball was the 
preferred ball there was not enough statistical evidence to suggest ball type was 
a statistically significant contributing factor to the success of making a shot.
Making a successful shot was greatly affected by distance as increased distance 
significantly decreased the chance of making a successful shot.  





### Introduction

Basketball requires shooting the ball in order to secure points for your team 
which will ultimately decide the outcome of the game. The results of the effects 
of distance on ball type will be an important contributing factor since shots can be made
from any distance within the court in a basketball game. Ball type is also an interesting 
variable as we can now investigate why the ball type basketball is exclusively 
used in the sport over other varying ball types and if its a significant contributing
factor in making a successful shot. This experiment will investigate the effects 
of  the ball types Soccerball and Netball compared to Basketball. The 2nd predictor
variable is the distance of the shot where shots were taken at three different distances;
1.5m, 3m, 4.5m.

This experiment will provide the ability to predict the effects of distance and differing
ball types on making a successful shot. This experiment will generate insight and 
essential information for aspiring basketball players as to where they may want to 
practice shooting for games or may lead them to take more in game shots at distances
with a higher chance of successfully making the shot. Once the experiment has 
concluded, some analysis will be conducted using a binomial distribution to examine
the effects of the 2 predictor variables previously described on making a successful
shot. This distribution will be conducted with a generalized linear model, a binomial
family as well as a link function of logit. This model will provide useful insight into 
the effects of distance and ball type on making a successful shot. 






### Experimental Design/ Observation (Including design and data collection method)


# aim 
This experiment aims to investigate the effects of different ball types at different 
distances has on making a successful shot in a basketball hoop.


# method 
This experiment will be undertaken at a local indoor leisure centre on indoor basketball
courts in order to reduce any existing natural variables ie wind resistance etc.
The experiment has 2 predictor variables which are distance and Ball type. For 
Ball type there are 3 different ball types used; basketball, Soccerball and
Netball. These 3 balls are the standard sizes used in games Basketball(size 7),
Soccerball(Size 5) and Netball(Size 5) weighing about 600 g, 450 g, and
450g respectively. The 2nd predictor variable is distance which will be three 
measurements made from the back of the rim using a tape measure to distinguish
where to mark each point with some duct tape. Once each shooting point is marked 
the three balls will be shot by one person from the smallest distance to the largest distance and
then repeated. At each distance the balls will be shot in a randomly selected order
to ultimately limit any potential effects of friction and heat within the ball so 
the results will maintain some validity. Each distance will be repeated 4 times
with 3 balls shot at each distance each time so we have a total of 36 observations (3*4*3).
Each time a shot was made it will be marked down as a miss (0) or make (1) and the corresponding
distance and ball type will be recorded with it. This gives rise to a binomial response
which will be analysed while utilising a generalized linear model with a logit link function

# Variables
Predictors:
Distance:
1.5 metres, 3 metres, 4.5 metres (All measured from the back of the rim of hoop)
Ball Type:
Basketball (size 7) 600g
Soccerball (size 5) 450 grams
Netball (size 5) 450 grams

Response:
The response variable is whether the shot goes through the hoop or not to see the 
effects of different ball types and distances on making a successful shot. Hence,
0 for missed shot, 1 for made shot



```{r include=FALSE}
shots <- read.csv("Shots.csv", header = TRUE, sep = ",")
show(shots)

shot1 <- factor(shots$shot)
shots$shot1 <- shot1
ball1 <- factor(shots$ball)  
shots$ball1 <- ball1
shots$ball <- NULL
shots$shot <- NULL
show(shots)

```


## Exploratory Data Analysis


```{r echo=FALSE}
library(ggplot2)

summary(shots)

plot(shot1~distance, data=shots, ylab = "Succesful shot")

plot(shot1~ball1, data=shots, xlab="Ball type", ylab = "Successful shot")

```
From the above plot of distance against making a successful shot we can observe a
decrease in successful shots as distance increases which could indicate a strong
negative linear relationship.

The plot of ball type against making a successful shot displays an even distribution 
of successful shots between netball and soccerball with the highest success rate 
being basketball. There is no obvious relationship visible here. 

## Statistical Analysis (Model selection process)

The response of making the basket in each case will be a binary response in the form of a
count as 0 or 1, where 0 is a missed shot and 1 is successful shot in the basket. 
This means the response will follow a binomial distribution. To model the binomial
distribution, a logistic regression will be used which is a member of the Generalized
Linear Model with a binomial response and contains a logistic link. The
formula for a logistic regression is:

$$ logit(\pi_{i}) = log(\frac{\pi_i}{1-\pi_i}) = \beta_0 + \beta_1 X_1 + \beta_2 X_2 +... +\beta_n X_n$$
Where 
$\pi_i$ is the probability of success (Making a shot)
$\beta_0$ is the intercept
$\beta_i$ is the coefficient for each variable, where i= 1:n
$X_{1:n}$ is each of the predictor variables


## Dummy Variables
In order to use categorical predictive variables in logistic regression a dummy 
variable will be created for Basketball (ball type) as this is the preferred ball type
this will be the dummy variable against soccer ball and netball to be able to see 
the difference in making a shot between this ball type and the other ball types.




## Logistic Regression Model
The first model was built using Generalised Linear Models with the family argument 
of binomial distribution and a link of logit. This model will contain the distance 
of the shot and the ball type without any interactions.

```{r echo=FALSE}
shots.glm <- glm(shot1 ~ ball1 + distance, family = binomial(link = "logit"), data = shots)
summary(shots.glm)
anova(shots.glm, test="Chisq")

```
## Hypothesis Test for Ball type and Successful Shots:

Null Hypothesis: $H_0:\beta_i = 0$
There is no difference between different ball type and making a shot.

Alternative Hypothesis: $H_A:\beta_i \neq 0$
There is at least one ball type which differs from the ball type basketball on 
making a shot.


Conclusion: 

From the above results we can observe a p-value of 0.639699 which is greater than
the 5% significance level and hence we fail to reject the null and we can conclude
that There is no statistically significant difference between different ball type and 
making a shot.

## Hypothesis Test for Distance and Successful shots:

Null Hypothesis: $H_0:\beta_1 = 0$
Distance does not have a statistically significant effect on successfully making a shot. 

Alternative Hypothesis: $H_A:\beta_1 \neq 0$
Distance does have a statistically significant effect on successfully making a shot. 



Conclusion: 

From the above results we can observe a p-value of 0.002681 which is less than
the 5% significance level. Therefore we reject the null and we can conclude
that distance has a statistically significant effect on successfully making a shot. 


## Model Explanation

Looking at the above results, it can be observed that ball types soccerball and netball
don't have statistically difference against ball type basketball in making a successful
shot as they both had a p-value greater than the 5% significance level. However since
these are included in the model as they are basketball types, a decrease of 0.9 in
the log odds of making a successful shot for both soccerball and netball.

For the results above we can also observe that distance of shot does have a 
statistically significant effect on making a successful shot as the p-value was 
less than the 5% significance level. Here we can observe that distance decreased 
the log odds of making a successful shot by -0.9280. From this model we will further
investigate whether we can produce a better model using the stepAIC() function.


# Model Formula

$$logit(Y_i) = 3.2263 + -0.8845X_1 -0.8845X_2 -0.9280X_3   $$
Where: $X_1 = Soccerball$, $X_2 = Netball$ and $X_3 = Distance$



# Stepwise AIC Approach for Model Selection:

```{r include = FALSE}
library(MASS)
shots_aic <- stepAIC(shots.glm,scope=list(lower="~1",upper="~( ball1 + distance)^2"))

```
```{r echo=FALSE}
summary(shots_aic)
anova(shots_aic, test = "Chisq")
```



### Logistic regression Model (Final Model Selected)

The next model will be constructed with the exclusion of ball type. This approach
was also used via the stepwise AIC function and the same model was produced as the 
best fit to the data.

```{r echo=FALSE}
shots.glm.d <- glm(shot1 ~ distance, family = binomial(link = "logit"), data = shots)
summary(shots.glm.d)
anova(shots.glm.d, test="Chisq")

```
## Hypothesis Test for Distance and Successful shots:

Null Hypothesis: $H_0:\beta_1 = 0$
Distance does not have a statistically significant effect on successfully making a shot. 

Alternative Hypothesis: $H_A:\beta_1 \neq 0$
Distance does have a statistically significant effect on successfully making a shot.

Conclusion:

From the above results we can observe a p-value of 0.003113 which is far less than
the 5% significance, hence we can reject the null hypothesis and conclude that 
distance has a statistically significant effect on making a successful shot.


## Diagnostic Plots

```{r}
par(mfrow=c(2,2))
plot(shots.glm.d)


```
# Residuals vs Fitted

The residual vs fitted plot shows an even number of points above and below the 
red line through the centre and the red line is flat, hence showing no heteroscedasticity.
A pattern is visible, although the residuals vs fitted plot will always show a 
pattern based on the nature of the success response.

# Normal QQ

From the QQ plot we can see the observations vary from the line and have longer tails 
than what were expected.

# Scale Location

There is a slight pattern in the Scale location plot although as discussed in the 
residual vs fitted plot patterns tend to occur in logistic regression plots due 
to the nature of the discrete counts of success in the model.

# Residuals vs Leverage

The observed observations are all seen to be within cooks distance.



## Effects plots

```{r}
library(effects)
plot(allEffects(shots.glm.d))


```
From the distance effects plot above we can observe that as shooting distance increased
the shooting success rate decreased.


### Discussion

From the analysis we saw that the 2nd model produced the smallest deviance and was
chosen by the Stepwise AIC approach. The 2nd model provided a deviance of about 8.7403,
whereas the first model provided a deviance of 9.0127. The diagnostic plots above
informs us that the model picked satisfies the conditions and can therefore be 
chosen to help describe and interpret the dataset produced from the experiment although
since our ball type variable wasn't statistically significant this model will be used 
only to explain the effects of distance on making a successful shot. 


### Final Model Explanation

The final model formula:

$$ logit(Y) =  2.5380 + -0.8940X $$ 
Where X is the predictor variable distance of shot.

Distance was always likely to have a greater effect on making a successful shot in
comparison to ball type as we saw from the results and chosen model. Although we 
did see from our exploratory analysis that Basketball had more shots made than the 
other 2 balls which were exactly equal in making successful shots. For further study 
this experiment could be undertaken by a professional Basketball player who has a 
higher shot accuracy than the non-athlete who conducted this experiment and is more 
familiar with shooting a basketball so we could possibly see some more interesting 
results and larger variation between differing ball types since someone shooting a ball who 
is relatively unskilled or under practiced in shooting a ball will not vary much from 
different ball types.




### Conclusion 

In conclusion, distance was proven to be the only statsticaly significant contributing 
factor in relation to making a successful shot. This is due to many factors such as 
more strength being required to keep the ball in the air for longer and with more strength
applied to the shot accuracy will typically be sacrificed especially by an experienced 
basketball shooter. The model produced with distance as the only predictor variable 
provided a lower AIC and deviance then the original model including ball type.
Overall the results produced from the final model can help generate insight and 
essential information for aspiring basketball players as to where they may want to 
practice shooting for games or may lead them to take more in game shots at distances
with a higher chance of successfully making the shot.


### References

Intro2r.info. 2022. Binomial data. [online] 
Available at: <https://www.intro2r.info/unit4/binomial-data.html> [Accessed 15 May 2022].

Timeless Basketball. 2022. The Importance of Sharp Shooting Skills — Timeless Basketball. 
[online] Available at: <https://timelessbasketball.com/basketball-shooting-skills/
#:~:text=Shooting%20is%20the%20most%20essential,all%20those%20skills%20to%20matter.> 
[Accessed 15 May 2022].

Eastly, A., 2022. STAT3030 Project. [report, PDF] p.16. Available at: <http://file:///C:/Users/edenf/Downloads/Example%20Report.pdf> [Accessed 15 May 2022].



























